---
title: "Ocean_Acidification_Lab"
author: "Irais Luquis Ramos"
date: "2023-03-23"
output: 
  prettydoc::html_pretty:
    theme: cayman
---

Can we quantify ocean acidification in the subtropical North Atlantic Ocean?
  
  1) Is surface ocean pco2 increasing
  2) Is surface ocean seawater ph decreasing? 
  3) Is surface ocean seawater saturation state with respect to aragonite           decreasing? 
  
Load Required library
```{r, message=FALSE}
library(tidyverse)
library(reader)
library(gsw)
library('seacarb')
```
BATS Bottle Import Data
```{r, message=FALSE, warning=FALSE}
bats_bottle <- read_delim("bats_bottle.txt", 
                          delim = "\t", escape_double = FALSE, 
                          col_names = FALSE, trim_ws = TRUE, skip = 60)

bats_bottle <- read_csv("bats_bottle.txt", 
                        skip = 59)

colnames(bats_bottle)<- colnames(read_csv("bats_bottle.txt", 
                                          skip = 59))
```

We are using BATS DATA 

Variables and Units 

yyyymmdd = Year Month Day   
decy   = Decimal Year     
time   = Time (hhmm)      
latN   = Latitude (Deg N) 
lonW   = Longitude (Deg W)
Depth  = Depth (m)                  
Temp   = Temperature ITS-90 (C)    
CTD_S  = CTD Salinity (PSS-78)      
Sal1   = Salinity-1 (PSS-78)        
Sig-th = Sigma-Theta (kg/m^3)       
O2(1)  = Oxygen-1 (umol/kg)          
OxFixT = Oxygen Fix Temp (C)        
Anom1  = Oxy Anomaly-1 (umol/kg)    
CO2    = dissolved inorganic carbon (umol/kg)              
Alk    = Alkalinity (uequiv)        
NO31   = Nitrate+Nitrite-1 (umol/kg)
NO21   = Nitrite-1 (umol/kg)        
PO41   = Phosphate-1 (umol/kg)      
Si1    = Silicate-1 (umol/kg)       
POC    = POC (ug/kg)                
PON    = PON (ug/kg)                
TOC    = TOC (umol/kg)                
TN     = TN (umol/kg)  
Bact   = Bacteria enumeration (cells*10^8/kg)   
POP    = POP (umol/kg)
TDP    = Total dissolved Phosphorus (nmol/kg)
SRP    = Low-level phosphorus (nmol/kg)
BSi    = Particulate biogenic silica (umol/kg)
LSi    = Particulate lithogenic silica  (umol/kg)
Pro    = Prochlorococcus (cells/ml)
Syn    = Synechococcus (cells/ml)
Piceu  = Picoeukaryotes (cells/ml)
Naneu  = Nanoeukaryotes (cells/ml)

  Quality flags
  
-999 = Missing or bad data
0 = Less than detection limit

##Intro on BATS and ocean acidification

Here we will be using The Gibbs SeaWater (GSW) Oceanographic Toolbox of TEOS-10 to helps us quantify ocean acidification over the past 40 years in the subtropical North Atlantic Ocean using BATS Time-series records. This with the intend to find different ways to predict changes going further into the future to help us understand what to expect in the future and analyze how this will affect the oceanic ecosystem. 